ODGI: understanding pangenome graphs

### 2021-10-18 manuscript/main.tex
Remove dot output
another potential discussion point
Intro: minor edits and comments

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}
\usepackage{subcaption}
\usepackage{algorithm2e}

@@

\newcommand{\vocab}{\textbf}

@@

A pangenome models the full set of genomic elements in a given species or clade \citep{cpang2018,Eizenga_2020}.
In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
These flexible models let us encode any kind of variation, allowing the generation of truly unbiased data systems on which to base analyses of genome evolution.
Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.

Projects like the Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to understand pangenomes.
As experience in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.

Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a genome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.

@@

@@

>>>>

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}
\usepackage{subcaption}
\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}

@@

\newcommand{\vocab}{\textbf}
\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}

@@

A pangenome models the full set of genomic elements in a given species or clade \citep{cpang2018,Eizenga_2020}.
In contrast to reference-based approaches which relate sequences to a single genome, these data structures encode the mutual relationships between all the genomes represented.
In pangenome graphs \citep{Paten:2017}, homologous regions between genomes are compressed into a single representative model of all alleles present in the pangenome.
These flexible models let us encode any kind of variation, allowing the generation of truly unbiased \FIXME{unbiased needs a little explanation} data systems on which to base analyses of genome evolution.
Although these structures are of clear utility to researchers \citep{cpang2018}, and have been the focus of numerous applications in read mapping \citep{Garrison:2018,Baaijens_2019,Hickey:2020,Sibbesen_2021}, we observe that the community lacks a toolset specifically focused on graph manipulation and interrogation.

The Human Pangenome Reference Consortium (HPRC) and Telomere-to-Telomere (T2T) consortium \citep{Miga:2020, Logsdon_2021, Nurk_2021} have recently demonstrated that high-quality \textit{de novo} assemblies can be routinely generated from third-generation long read sequencing data.
We anticipate that in the near future, studies will frequently begin from assemblies of similar quality, leading to demand for methods that allow us to create and explore pangenomes.
As experience \FIXME{this is a bit odd} in the HPRC has shown, pangenome graphs are a natural model to support these kinds of analyses, but studies based on pangenome graphs are difficult due to a lack of downstream tools and standard operations.

Here, we present the Optimized Dynamic Genome/Graph Implementation (ODGI) toolkit, a pangenome graph interrogation and transformation system specifically built to handle the data scales encountered when working with pangenomes built from hundreds of haplotype-resolved genomes.
ODGI provides a set of standard operations on the pangenome variation graph data model, generalizing ``genome arithmetic'' concepts like those found in BEDTools \citep{Quinlan_2010} to work on genome graphs, and providing new operations, such as visualizations, sorting, and liftover projections critical to understanding and exploiting pangenome graphs.

@@

\texttt{odgi view} can output GFA and internal graph representations \\
\texttt{odgi stats} outputs statistics in TSV or YAML. \\
\texttt{odgi bin} can output JSON. \\
\texttt{odgi degree/position/depth/untangle/tips} can output BED. \\
\texttt{odgi untangle} outputs PAF.

@@

is ODGI ready for scWGS data? potentially, each cell is a path in a graph!! very very high path depth
